Poker Chips, Earthquakes, and Computation

### 2014-10-13 finaldraft.tex
edited some minor things
added guiding text to d trees section

\begin{corollary}
\label{norepeats}
A finite board can never return to a previous state via a sequence of earthquakes.
\end{corollary}
\begin{proof}
Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$. Lemma~\ref{spreadfunction} shows that $\psi(b) < \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
\end{proof}

@@

Now that we have shown that all finite boards will eventually reach a stable state we would like to know if the order in which these earthquakes occur matters. More specifically, if we consider some finite starting board and two sequences of earthquakes that stabilizes the board, we would like to know if the resultant board is the same for both sequences.

@@

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$ if $E$ is a valid earthquake, then 
\begin{equation*}
E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state.
\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2 E_1 b = E_1 E_2 b
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

We demonstrate this by contradiction. Assume  $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$ we know $b(s_1) \geq 4 $. 
Therefore we must have $b(s_1)<8$ and $s_2=s_1$. But then $E_1 b(s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes. This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.

Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2 E_1 b(x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$. Applying Lemma ~\ref{earthquakeredefine} we have 
\begin{align*}
E_2 E_1 b (x) = E_1(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
= b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
= b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
= E_2 b(x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
= E_1 E_2 b(x) 
\end{align*}
The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
\end{proof}

We can use this lemma along with a simple induction to show that under appropriate conditions we can move an earthquake earlier in the sequence without affecting the final board state.
\begin{lemma}
Let $E_1,E_2,E_3,\dots, E_n$ be a sequence of earthquakes, where $E_k$ hits the node $s_k$. Consider  the sequence given by moving some earthquake $E_j$ to position $m<j$ in the sequence, such that $E_j$ is a valid earthquake for all positions between $m$ and $j$, inclusive. 
This sequence is valid and produces the same final board configuration.
\end{lemma}
\label{shiftlemma}
\begin{proof}
It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes. In fact we can also ignore the first $m-1$ terms, because their order is not changed so the board state will be identical after $m-1$ terms for both sequences. Let us call this board state $b_{m-1}$

@@

We can show this by inducting on the length of the sequence $l=j-m+1$. The base case of $l=2$ is shown in lemma ~\ref{swappinglemma}. Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$. So our sequence is 
\begin{equation*}
E_m, E_{m+1}, \ldots, E_{j-1}, E_j
\end{equation*}
where $p+1=j-m+1$. Then if we set $b_{m} = E_{m} (b_{m-1})$ by applying our inductive step we have that the two sequences 
\begin{align*}
E_{m+1}, \ldots, E_{j-1}, E_{j} \\ 
E_{j}, E_{m+1}, \ldots,  E_{j-1} 
\end{align*}
are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma ~\ref{swappinglemma} we know that the two sequences 
\begin{align*}
E_{m}, E_{j} \\
E_{j}, E_{m}
\end{align*}
are both valid and produce the same final board configuration. If we apply the same reduction we used in our first step of this proof we have  that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$  and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board. 
\end{proof}

Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board we can show that the number of earthquakes hitting a square for a given board is the same regardless of order. 
The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen, based on the initial board and then move these earthquakes to the front of the sequence without changing the final board. This technique will specify an order in which the earthquakes will occur, depending only on the original board which completes the proof.

@@

Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the nodes. We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes. We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of Lemma ~\ref{shiftlemma}).

@@

\begin{definition}
A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each are connected by edges to $d-1$ unique children and so on.  
\end{definition}
%\begin{figure}[!h]
%\label{4tree}
%\centering
%\includegraphics[width=0.4\linewidth]{4tree}
%\caption{A 4-tree}
%\end{figure}
We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
\begin{definition}
A joined $d$-tree is two $d$-trees connected at their parent nodes by an edge.
\end{definition}
%\begin{figure}[!h]
%\label{joined4tree}
%\centering
%\includegraphics[width=0.8\linewidth]{joined4tree}
%\caption{A joined 4-tree}
%\end{figure}

We now define a coordinate system so that our method of proof can mirror that of the standard board case.
\begin{definition} We define a coordinate system for our joined $d$-trees, with the following rules:
\begin{enumerate}
	\item $(0,0)$ and $(1,0)$ are valid coordinates and refer to the two parent nodes.
	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq (d-1)^{|a|}$
	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq (d-1)^{|a-1|}$
\end{enumerate}
We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
\begin{equation*}
G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
\end{equation*}
\end{definition}

>>>>

\begin{corollary}
\label{norepeats}
A finite board can never return to a previous state via a sequence of earthquakes.
\end{corollary}
\begin{proof}
Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$. Lemma~\ref{spreadfunction} shows that $\Psi(b) < \Psi(E_1E_2\cdots E_kb)$, which is a contradiction.
\end{proof}

@@

Now that we have shown that all finite boards will eventually reach a stable state we would like to know if the order in which these earthquakes occur matters. More specifically, if we consider some finite starting board and two sequences of earthquakes that stabilize the board, we would like to know if the resultant board is the same for both sequences.

@@

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$. If $E$ is a valid earthquake, then 
\begin{equation*}
E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

Now we can show that switching the order of two earthquakes with certain validity conditions does not affect the final board state. Note that we will frequently refer to sequences of earthquakes. Earthquake sequences correspond to those earthquakes being applied in the order the sequence specifies. For example, if we have a sequence of earthquakes $E_1,E_2,E_3$ applied to board $b$, the resulting board is $E_3E_2E_1b$.

\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2 E_1 b = E_1 E_2 b
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

We prove this by contradiction. Assume $E_2, E_1$ is an invalid sequence, that is $E_2 b(s_1)<4$. Since $E_1$ is a valid earthquake when applied directly to $b$, we know $b(s_1) \geq 4 $. 
Therefore we must have $b(s_1)<8$ and $s_2=s_1$. But then $E_1 b(s_1)<4$ and so $E_1, E_2$ would not be a valid sequence of earthquakes. This contradicts one our hypothesis and so $E_2, E_1$ is a valid sequence.

Now in order to demonstrate that the order in which the earthquakes are applied does not affect the final board state consider $E_2 E_1 b(x)$ for some $x \in \mathbb{Z} \times \mathbb{Z}$. Applying Lemma~\ref{earthquakeredefine} we have 
\begin{align*}
E_2 E_1 b (x) = E_1(x)  + [[ x \in n(s_2) ]] - 4[[ x=s_2]] \\
= b(x) + [[ x \in n(s_1) ]] - 4[[x = s_1] + [[ x \in n(s_2) ]] - 4[[ x= s_2)]]  \\
= b(x) + [[ x \in n(s_2) ]] - 4[[x = s_2] + [[ x \in n(s_1) ]] - 4[[ x= s_1)]]\\
= E_2 b(x) + [[ x \in n(s_1) ]] - 4[[ x = s_1]] \\
= E_1 E_2 b(x) 
\end{align*}
The critical step in this proof is recognizing that because $E_1$ and $E_2$ are valid earthquakes regardless of their ordering, the algebra above is valid.
\end{proof}

We can use this lemma along with a simple induction to show that under appropriate conditions we can move an earthquake to an earlier position in the sequence without affecting the final board state.
\begin{lemma}
Let $E_1,E_2,E_3,\dots, E_n$ be a sequence of earthquakes, where $E_k$ hits the node $s_k$. Consider  the sequence given by moving some earthquake $E_j$ to position $m<j$ in the sequence, such that $E_j$ is a valid earthquake for all positions between $m$ and $j$, inclusive. 
This sequence is valid and produces the same final board configuration.
\end{lemma}
\label{shiftlemma}
\begin{proof}
It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes. In fact we can also ignore the first $m-1$ terms, because their order is not changed so the board state will be identical after $m-1$ terms for both sequences. Let us call this board state $b_{m-1}$

@@

We can show this by inducting on the length of the sequence $l=j-m+1$. The base case of $l=2$ is shown in Lemma~\ref{swappinglemma}. Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$. So our sequence is 
\begin{equation*}
E_m, E_{m+1}, \ldots, E_{j-1}, E_j
\end{equation*}
where $p+1=j-m+1$. Then if we set $b_{m} = E_{m} (b_{m-1})$ by applying our inductive step we have that the two sequences 
\begin{align*}
E_{m+1}, \ldots, E_{j-1}, E_{j} \\ 
E_{j}, E_{m+1}, \ldots,  E_{j-1} 
\end{align*}
are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma~\ref{swappinglemma} we know that the two sequences 
\begin{align*}
E_{m}, E_{j} \\
E_{j}, E_{m}
\end{align*}
are both valid and produce the same final board configuration. If we apply the same reduction we used in our first step of this proof we have that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board. 
\end{proof}

Now that we have developed a tool that will allow us to move earthquakes in a sequence without changing the final board, we can show that the number of earthquakes hitting a square for a given board is the same regardless of order.

The general idea is that we can recursively specify a sequence of earthquakes that ``must" happen based on the initial board. We can then move these earthquakes to the front of the sequence without changing the final board. This technique will specify an order in which the earthquakes will occur, depending only on the original board, which completes the proof.

@@

Now, we find the end of the sequence of moved earthquakes and consider the board at this point, noting that we have new values for all the nodes. We repeat this argument, moving a new set of earthquakes directly to the right of previously moved earthquakes. We repeat this until the end, and we note that eventually this argument terminates (because if we get to a point where there are still more earthquakes ahead, and the board doesn't allow any valid earthquakes, this is a contradiction of Lemma~\ref{shiftlemma}).

@@

So far, we have analyzed the poker chips problem assuming that the underlying graph is an infinite grid graph. One important feature of this graph is that it is \emph{4-regular}, which means that each node has exactly 4 neighbors. Graphs where every node has an equal number of neighbors allow for an easy definition of the earthquake operation; if every node has $d$ neighbors, then an earthquake redistributes $d$ chips to each of the $d$ neighbors.

Thus, in this section we analyze how the problem changes when played over a large class of graphs that are regular, but are not grid-shaped.

\begin{definition}
A $d$-tree is a graph composed of a parent node connected by edges to $d-1$ unique children nodes.Those children each are connected by edges to $d-1$ unique children and so on.  
\end{definition}
\begin{figure}[!h]
\label{4tree}
\centering
\includegraphics[width=0.4\linewidth]{4tree}
\caption{A 4-tree}
\end{figure}
We note that every node other than the original parent node has $d$ neighbors. In order to simplify our process we would like the graph to be regular and so we make the following definition. 
\begin{definition}
A joined $d$-tree is two $d$-trees connected at their parent nodes by an edge.
\end{definition}
\begin{figure}[!h]
\label{joined4tree}
\centering
\includegraphics[width=0.8\linewidth]{joined4tree}
\caption{A joined 4-tree}
\end{figure}

We now define a coordinate system so that our method of proof can mirror that of the standard board case.
\begin{definition} We define a coordinate system for our joined $d$-trees, with the following rules:
\begin{enumerate}
	\item $(0,0)$ and $(1,0)$ are valid coordinates and refer to the two parent nodes.
	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq (d-1)^{|a|}$
	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq (d-1)^{|a-1|}$
\end{enumerate}
We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
\end{definition}
\begin{equation*}
G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
\end{equation*}
